---
title: "CodeBook"
author: "Kevin Costello"
date: "04/05/2017"
output: 
        html_document:
                keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, echo = FALSE, results = "markdown")
```

```{r libraries}
# Load libraries needed for assignment
library(dplyr)
library(data.table)
library(tidyr)
library(stringr)
library(xtable)
```

### 1. Print out of raw text files read into R.

```{r input text files}
# Set working directory for assignment
setwd("~/Coursera/Getting and Cleaning Data/Week 4/UCI HAR Dataset")

# Create function to read in text files and output to a data frame
inputted <- function(input_file, output_file, columns, headObs) {
  file = paste(getwd(),input_file, sep = "")
  output_file <- read.table(file = file, sep="", stringsAsFactors=FALSE, col.names = columns)
  inputName <- deparse(substitute(input_file))
  phone <- "[[:alpha:]_]+[\\.]+txt"
  extractChar <- str_extract(inputName, phone)
  #print(inputName)
  print(str_c("Filename: ",extractChar[!is.na(extractChar)]))
  cat("\n")  
  print(head(output_file, n = headObs))  
  return(output_file)
}

# Input test data
y_test <- inputted("/test/Y_test.txt",y_test, c("activityCode"), 2)
activity_labels <- inputted("/activity_labels.txt",activity_labels, c("activityCode","activityLabel"), 6)
features <- inputted("/features.txt",features, c("measurementCode","measurementLabel"), 561)
x_test <- inputted("/test/X_test.txt",x_test, paste0("V", seq_len(length(features$measurementCode))), 2)
subject_test <- inputted("/test/subject_test.txt",features, c("subjectID"), 2)

# Input training data
x_train <- inputted("/train/X_train.txt",x_train, paste0("V", seq_len(length(features$measurementCode))), 2)
y_train <- inputted("/train/Y_train.txt",y_train, c("activityCode"), 2)
subject_train <- inputted("/train/subject_train.txt",features, c("subjectID"), 2)
```
***

### 2. Process the test data files.

```{r process test data}
# Combine y_test and x_test by row
test_all <- cbind(subject_test, y_test, x_test)

# Merge data and activity labels by activityCode value
test_all_labels <- left_join(test_all, activity_labels, by = "activityCode")

print("Print first two rows of merged test dataset (dataset test_all_labels)")
head(test_all_labels, n = 2)

# Transpose activity labels from vertical to horizontal structure
trans_features <- transpose(features)

# Select just the names column from trans_features
trans2 <- trans_features[2,]

# Convert list of names to vector
trans3 <- as.vector(trans2, mode = "character")

# Rename variables from x_test using the labels from the features file
test_all_labels_copy <- test_all_labels
searchTerms <- "V"
variablesToKeep <- str_detect(names(test_all_labels_copy), searchTerms)
names(test_all_labels_copy)[variablesToKeep] <- trans3

print("Print the transformed variable names")
print(cbind(names(test_all_labels)[variablesToKeep], names(test_all_labels_copy)[variablesToKeep]))

# Keep columns which contain mean or standard deviation statistics
searchTerms <- "std\\(|mean\\(|activityCode|activityLabel|subjectID"
variablesToKeep <- str_detect(names(test_all_labels_copy), searchTerms)
test_mean_sd <- test_all_labels_copy[,variablesToKeep] 
test_final <- test_mean_sd %>%
        select(subjectID, activityLabel, activityCode, starts_with("t"), starts_with("f"))

print("Print the first two rows of the final test dataset (dataset test_final)")
head(test_final, n = 2)
```

### 3. Process the train data files.

```{r process train data}
# Combine y_train and x_train by row
train_all <- cbind(subject_train, y_train, x_train)

# Merge data and activity labels by activityCode value
train_all_labels <- left_join(train_all, activity_labels, by = "activityCode")

print("Print first two rows of merged train dataset (dataset train_all_labels)")
head(train_all_labels, n = 2)

# Rename variables from x_train using the labels from the features file
train_all_labels_copy <- train_all_labels
searchTerms <- "V"
variablesToKeep <- str_detect(names(train_all_labels_copy), searchTerms)
names(train_all_labels_copy)[variablesToKeep] <- trans3

print("Print the transformed variable names")
print(cbind(names(train_all_labels)[variablesToKeep], names(train_all_labels_copy)[variablesToKeep]))

# Keep columns which contain mean or standard deviation statistics
searchTerms <- "std\\(|mean\\(|activityCode|activityLabel|subjectID"
variablesToKeep <- str_detect(names(train_all_labels_copy), searchTerms)
train_mean_sd <- train_all_labels_copy[,variablesToKeep]
train_final <- train_mean_sd %>%
        select(subjectID, activityLabel, activityCode, starts_with("t"), starts_with("f"))

print("Print the first two rows of the final train dataset (dataset train_final)")
head(train_final, n = 2)

# Merge measurement code values from features file
searchTerms <- "std\\(|mean\\("
variablesToKeep <- str_detect(names(train_all_labels_copy), searchTerms)
measurementLabel <- names(train_all_labels_copy)[variablesToKeep]
measurementFrame <- as.data.frame(measurementLabel, stringsAsFactors = FALSE)
measurementFeatures <- right_join(features, measurementFrame, by = "measurementLabel")
```

### 4. Merge the test and train datasets.

```{r Merge test and train datasets}
# Combine test and train datasets by row
test_train_final <- rbind(test_final, train_final)
print("Print the first two rows of the final dataset in wide form (dataset test_train_final)")
head(test_train_final, n = 2)
```

### 5. Restructure the merged datasets from wide to narrow format. This means that only one variable (Variable value) is needed to contain the values for each measurement.

|Dataset finalDataset description|
|--------------------------------|

|Variable name           |Description                                                    |
|------------------------|---------------------------------------------------------------|
|measurementLabel        |Measurement name taken from features.txt file                     |
|measurementCode         |Measurement code taken from features.txt file                     |
|subjectID               |Subject ID taken from subject_test.txt and subject_train.txt   |
|activityLabel           |Activity name taken from activity_labels.txt file              |
|activityCode            |Activity code taken from activity_labels.txt file              |
|value                   |Value for each measurement taken from x_test.txt and x_train.txt|

|Print first 100 rows of Dataset finalDataset in long/narrow form|
|----------------------------------------------------------------|

```{r restructure data}
# Restructure data from wide to narrow format
# This means fewer variables are needed to contain the mean value for each variable
narrow <- test_train_final %>%
        gather(key = measurementLabel, value = value, `tBodyAcc-mean()-X`:`fBodyBodyGyroJerkMag-std()`, factor_key = FALSE) %>%
        select(measurementLabel, subjectID, activityCode, activityLabel, value)

finalDataset <- left_join(narrow, measurementFeatures, by ="measurementLabel") %>%
        select(measurementLabel, measurementCode, subjectID, activityLabel, activityCode, value) %>%
        arrange(measurementCode, subjectID, activityCode)
        
head(finalDataset, n = 100)
```


|Print list of all subjects in Dataset finalDataset|
|--------------------------------------------------|

```{r printAllSubjects}
print(distinct(finalDataset, subjectID))
```

|Print list of all measurements in Dataset finalDataset|
|------------------------------------------------------|

```{r printAllMeasurements}
print(distinct(finalDataset, measurementLabel, measurementCode))
```

|Print list of all activities in Dataset finalDataset|
|----------------------------------------------------|

```{r printAllActivities}
print(distinct(finalDataset, activityLabel, activityCode))
```

### 6. Calculate the mean of each measurement by subject and activity.

|Dataset meanDataset description|
|-------------------------------|

|Variable name                           |Description                                    |
|----------------------------------------|-----------------------------------------------|
|measurementLabel                        |Measurement name taken from features.txt file     |
|measurementCode                         |Measurement code taken from features.txt file     |
|subjectID                               |Subject ID taken from subject_test.txt and      subject_train.txt                                                                        |
|activityLabel                           |Activity name taken from activity_labels.txt  file                                                                                     |
|activityCode                            |Activity code taken from activity_labels.txt file                                                                                     |
|meanMeasurementBySubjectByActivity      |Mean value of each measurement by subject and activity     |

|Print first sixty rows of Dataset meanDataset|
|---------------------------------------------|

```{r calculate mean of each variable by subject and activity}
# Calculate mean for each variable by subject and activity
meanDataset <- finalDataset %>%
        group_by(measurementLabel, measurementCode, subjectID, activityLabel, activityCode) %>%
        summarise(meanMeasurementBySubjectByActivity = signif(mean(value), digits = 3)) %>%
        select(measurementLabel, measurementCode, subjectID, activityLabel, activityCode, meanMeasurementBySubjectByActivity) %>%
        arrange(measurementCode, subjectID, activityCode) %>%
        as.data.frame(.)
head(meanDataset, n = 60)
```

|Print list of all subjects in Dataset meanDataset|
|--------------------------------------------------|

```{r printMeanSubjects}
print(distinct(meanDataset, subjectID))
```

|Print list of all measurements in Dataset meanDataset|
|------------------------------------------------------|

```{r printMeanMeasurements}
print(distinct(meanDataset, measurementLabel, measurementCode))
```

|Print list of all activities in Dataset meanDataset|
|----------------------------------------------------|

```{r printMeanActivities}
print(distinct(meanDataset, activityLabel, activityCode))
```

### 7. Export the dataset containing the mean calculations to a text file. Include check to import the exported text file to R and compare with the dataset used to create the text file to ensure they are the same.

```{r export meanDataset to text file}
# Write tidy dataset of averages to text file
# Set working directory for text file
setwd("~/Coursera/Getting and Cleaning Data/Week 4/Peer Graded Assignment")
write.table(meanDataset, file = "tidy.txt", row.names = FALSE)

# Check to see that exported text file is the same as the dataset exported
indata <- read.table("tidy.txt", header = TRUE, stringsAsFactors = FALSE)

print("Compare exported text file with dataset exported")
identical(meanDataset, indata)
```
